---
title: "The partitions package: enumeration in R"
output:
  github_document:
    pandoc_args: --webtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# partitions

<!-- badges: start -->
[![Build Status](https://travis-ci.org/RobinHankin/partitions.svg?branch=master)](https://travis-ci.org/RobinHankin/partitions)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/partitions)](https://cran.r-project.org/package=partitions)
[![Rdoc](http://www.rdocumentation.org/badges/version/partitions)](http://www.rdocumentation.org/packages/partitions)

<!-- badges: end -->

# Overview

The ```partitions``` package provides efficient vectorized code to
enumerate solutions to various integer equations.  For example, we
might note that

$$
5 = 4+1 = 3+2 = 3+1+1 = 2+2+1 = 2+1+1+1 = 1+1+1+1+1
$$

and we might want to list all seven in a consistent format (note here
that each sum is written in nonincreasing order, so $3+1$ is
considered to be the same as $1+3$).

# Installation

You can install the released version of wedge from [CRAN](https://CRAN.R-project.org) with:

```{r, message=FALSE}
# install.packages("partitions")  # uncomment this to install the package
library("partitions")
```

# The ```partitions``` package in use

To enumerate the partitions of 5:


```{r}
parts(5)
```

(each column is padded with zeros).  Of course, larger integers have
many more partitions and in this case we can use ```summary()```:


```{r}
summary(parts(16))
```

Sometimes we want to find the unequal partitions (that is, partitions
without repeats):

```{r}
summary(diffparts(16))
```


## Restricted partitions

Sometimes we have restrictions on the partition.  For example, to
enumerate the partitions of 9 into 5 parts we would use
```restrictedparts()```:

```{r}
summary(restrictedparts(9,5))
```

and if we want the partitions of 9 into parts not exceeding 5 we would
use the conjugate of this:

```{r}
summary(conjugate(restrictedparts(9,5)))
```


## Block parts

Sometimes we have restrictions on each element of a partition and in
this case we would use ```blockparts()```:

```{r}
summary(blockparts(1:6,10))
```

which would show all solutions to $\sum_{i=1}^6a_i=9$, $a_i\leq i$.


## Compositions

Above we considered $3+2$ and $2+3$ to be the same partition, but if
these are considered to be distinct, we need the _compositions_, not
partitions:


```{r}
compositions(4)
```

## Set partitions

A set of 4 elements, WLOG $\{1,2,3,4\}$, may be partitioned into
subsets in a number of ways and these are enumerated with the
```setparts()``` function:

```{r}
setparts(4)
```

In the above, column ```2 3 1 1``` would correspond to the set
partition $\{\{3,4\},\{1\},\{2\}\}$.


# Further information
For more detail, see the package vignettes

```
vignette("partitionspaper")
vignette("setpartitions")
vignette("scrabble")
``` 
