---
title: "The partitions package: enumeration in R"
output:
  github_document:
    pandoc_args: --webtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# partitions

<!-- badges: start -->
[![Build Status](https://travis-ci.org/RobinHankin/partitions.svg?branch=master)](https://travis-ci.org/RobinHankin/partitions)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/partitions)](https://cran.r-project.org/package=partitions)
[![Rdoc](http://www.rdocumentation.org/badges/version/partitions)](http://www.rdocumentation.org/packages/partitions)

<!-- badges: end -->

# Overview

The ```partitions``` package provides efficient vectorized code to
enumerate solutions to various integer equations.  For example, we
might note that

$$
5 = 4+1 = 3+2 = 3+1+1 = 2+2+1 = 2+1+1+1 = 1+1+1+1+1
$$

and we might want to list all seven in a consistent format.



# Installation

You can install the released version of wedge from [CRAN](https://CRAN.R-project.org) with:

```{r, message=FALSE}
# install.packages("partitions")  # uncomment this to install the package
library("partitions")
```

# The ```partitions``` package in use

To enumerate the partitions of 5:


```{r}
parts(5)
```

(each column is padded with zeros).  Of course, larger integers have
many more partitions and in this case we can use ```summary()```:


```{r}
summary(parts(16))
```

Sometimes we want to find the unequal partitions (that is, partitions
without repeats):

```{r}
summary(diffparts(16))
```

