---
title: "A generating function for restricted partitions"
author: "Robin KS Hankin"
date: "23/09/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract 


A generating function for restricted partitions---originally due, as
far as I can tell, to Wilf (2000)---is presented and R idiom using the
`spray` package given.  The generating function approach is shown to
be not particularly efficient compared to the direct enumeration used
in
the `partitions` package.  

## Intoduction

The `partitions` package gives functionality for various integer
partition enumeration problems including that of restricted
partitions, function `restrictedparts()`:

```{r, label=userestrictedpartsfunction}
library("partitions")
(jj <- restrictedparts(7,3))
ncol(jj)
```

Here I will consider function `R()`, which calculates the size of
the matrix required:

```{r,label=useRopenclose}
R(3,7,include.zero=TRUE)
```

Function `R()` is very basic; all it does is to go through all the
restricted partitions, counting them one by one until the recursion
bottoms out:

```
unsigned int numbrestrictedparts(int *x, const int m){
        unsigned int count=1;

        while(c_nextrestrictedpart(x, &m)==0){
           count++;
        }
        return count;
```

To implement a potentially more efficient method, we can use
generating functions.  Here we follow Wilf and, using his terminology,
define an infinite polynomial $P(x,y)$ as follows:

\begin{equation}\label{GF}
P(x,y) = \prod_{r=0}^\infty\frac{1}{1-x^ry}
\end{equation}

Or, expanding:

\begin{equation}\label{GF2}
P(x,y) =
\left(1+y+y^2+y^3 + \cdots\right)
\left(1 + xy + x^2y^2 + x^3y^3+\cdots\right)
%\left(1 + x^2y + x^4y^2+x^6y^3+\cdots\right)
\cdots
\left(1+x^ry+x^{2r}y^{2}+x^{3r}y^{3}+\cdots\right)\cdots
\end{equation}

The power of $x$ counts the total of the chosen integers (the size of
the partition), and the power of $y$ counts the number of integers
chosen (the length of the partition).  Thus the number of partitions
of $k$ into at most $n$ parts is the coefficient of $x^ky^n$ in
$P(x,y)$.  In numerical work it is convenient and efficient to ignore
terms with a power of $x$ higher than $n$ (sum of integers chosen
exceeds $n$), or with power of $y$ higher than $k$ (number of integers
chosen exceeds $k$).  Taking `R(3,7,include.zero=TRUE)` as an example
we would truncate equation \ref{GF2} as follows:

\begin{eqnarray}\label{GF3}
P(x,y) = &{}&
\left(1+y+y^2+y^3\right)
\left(1 + xy + x^2y^2 + x^3y^3\right)
\left(1 + x^2y + x^4y^2+x^6y^3\right)\times\nonumber\\
&{}&\left(1 + x^3y + x^6y^2\right)
\left(1 + x^4y\right)
\left(1 + x^5y\right)
\left(1 + x^6y\right)
\left(1 + x^7y\right)
\end{eqnarray}

and the coefficients of $P(x,y)$ up to $x^7y^3$ would correctly count
the restricted partitions.  Note that we need consider only at most
four terms in each bracket (powers of $y$ above three being
irrelevant) and we may stop the continued product at the $x^7$ term as
further brackets contain only one and powers of $x$ above the eighth.

The R implementation uses the `spray` package, in particular
function `ooom(x)` which returns $\frac{1}{1-x}$.

```{r}
library("spray")
R_gf <- function(k,n){   # version 1
   x <- spray(cbind(1,0))
   y <- spray(cbind(0,1))
   P <- ooom(y,k)  # term x^0; number of zeros chosen
   for(i in seq_len(k)){  # starts at 1
     P <- P*ooom(x^i*y,n)
   }
   return(value(P[k,n]))
}
```


Thus

```{r}
R_gf(7,3)
```

We can do slightly better in terms of efficiency by ruthlessly cutting
out powers higher than needed:                
                         
```{r,label=define_strip}
strip <- function(P,k,n){  # strips out powers higher than needed
  ind <- index(P)
  val <- value(P)
  wanted <- (ind[,1] <= k) & (ind[,2] <= n)
  spray(ind[wanted,],val[wanted])
}
```

which is used here:
                
```{r, label=define_Rgf_version2}
R_gf2 <- function(k,n,give_poly=FALSE){
   x <- spray(cbind(x=1,y=0))
   y <- spray(cbind(x=0,y=1))
   P <- ooom(y,k)  # term x^0
   for(i in seq_len(k)){  # starts at 1
     P <- strip(P*ooom(spray(cbind(i,0))*y, min(n,ceiling(k/i))),k,n)
   }
   if(give_poly){
     return(P)
   } else {
     return(value(P[k,n]))
   }
}
```

then

```{r,gf2_use}
R_gf2(7,3)
```

#  Computational efficiency

We can test the computational efficiency of the generating function
approach using larger values of $k$ and $n$:


```{r}
k <- 140
n <- 4

system.time(jj1 <- R(n,k,include.zero=TRUE))
system.time(jj2 <- R_gf2(k,n))
jj1==jj2
```

So the generating function approach is not particularly efficient, at
least not in this sort of use-case with the `spray` package.  It
might be better with the `mvp` package; I don't know.

Of course, `R_gf2()}` calculates the generating polynomial
which gives very much more information than is returned.  Perhaps this
is why it is so slow compared to function `R()`, although it is
surprising to see direct enumeration so heavily outperforming a
generating function.


### References

H. S. Wilf, 2000. "Lectures on Integer Partitions"
